
# Part I:  Adjustment for Confounders  

How do we adjust for confounders?  The most common approach involves
[There will also be the question of *which* possible confounders to use.
]{.column-margin} linear models, with which we express the mean Y for
given values of the X, C and S variables in a linear form.  

::: {.callout-note}
### Important term: the *regression function*
The relation of mean Y to the X, C and S variables is formally called
the *regression function* of Y on those variables. Our current model
will assume this relation as linear, but the term *regression function*
is general.
:::

## Linear model example: a simple gender wage gap analysis

Consider the **svcensus** data example in @sec-census above,
investigating a possible gender pay gap.  So Y is wage and S is gender.
We might treat age as a confounder C, reasoning as follows.  Older
workers tend to have more experience and thus higher wages, and if there
is an age differential in our data, say with female workers tending to
be older, this may mask a gender pay gap.  

So, let's take the set of confounders C to consist of age, and for
simplicity in this introductory example, not include any other
confounders, such as occupation, and not include any other
variables X.

### Initial analysis

Our linear model would thus be

> mean W = $\beta$~0~ + $\beta$~1~ A + $\beta$~2~ M

[The column svcensus$gender is an R factor. Our function **dsldLinear**
calls R's **lm**, which replaces that column by a dummy variable
**gendermale**, our M above.]{.column-margin} where W is wage, A is age
and M is an indicator variable, with M = 1 for men and M = 0 for women.
The parameters $\beta$~i~ are estimated by fitting the model to the
data:

```{r}
svcensus1 <- 
   svcensus[,c(1,4,6)]  # age, wage, gender
z <- dsldLinear(svcensus1,'wageinc','gender')
coef(z)  # print the estimated coefficients b_i 
```

Let's use b~i~ to denote our estimated $\beta$~i~.  So for instance b~1~
[Always keep in mind that statistical quantities are only estimated,
since we work only with sample data from some population, real or
conceptual.  Hence the need for standard errors, confidence intervals
and so on.]{.column-margin} 
= 489.5728 is our estimate of the unknown population parameter
$\beta_1$.

### Interpretation of $\beta$~2~

Lots in the output to discuss, which we will gradually cover below.  For now, note that the estimate b~2~ turns out to be about $13,000, which is the (estimated) wage gap, if any.  Here's why:

Under the model, the mean wage for, say, 36-year-old men is

> $\beta$~0~ + 36 $\beta$~1~ + 1 $\beta$~2~ 

while for women of that age it is

> $\beta$~0~ + 36 $\beta$~1~ 

The difference is $\beta$~2~.  But if we look at, for instance, people of age
43, the mean wages for men and women are

> $\beta$~0~ + 43 $\beta$~1~ + 1 $\beta$~2~ 

and

> $\beta$~0~ + 43 $\beta$~1~ 

and the difference *is still* $\beta$~2~.  

::: {.callout-note}
### "The" effect of gender
Thus we can speak of $\beta$~2~ as *the* gender wage gap, at any age.
According to the model, younger men earn an estimated $13,000 more than
younger women, with the *same-sized* gap between older men and older
women.
:::

The above approach to dealing with confounders is a very common one.
But it raises questions, such as:

* What are the assumptions underlying that model?  And how might we [In
  addition, the data here are, as is commonly the case, *observational*,
  as opposed to being the result of a *randomized clinical trial*; there
  may be serious issues, due to unobserved confounders.  Such problems
  might be solvable via an advanced (and rather controversial)
  methodology known as *causal inference*.  Unfortunately, details are
  beyond our scope in this tutorial, but we will explain some basic
  concepts in @sec-causal.]{.column-margin} check whether they are
  (approximately) valid?

* We chose only one C variable here, age.  We might also include
occupation, as noted earlier.  In some datasets, might have dozens of
possible confounders.  How do we choose which ones to use in our model?
And for that matter, why not use them all?

* The above model, in which the gender wage gap was uniform across all
wages, may not be adequate.  How can we determine this, and what
alternative models might we use?

### Statistical inference

The full output of **dsldLinear()** goes to the heart of discrimination
analysis, enabling statistical inferences on differences in levels of
the sensitive variable S.  Let's take a look, continuing from the above
code: 

```{r}
summary(z)
```

The first half of this output is from **lm()**, which is called by
**dsldLinear()**.  The second half is the "value added" material 
from **dsld**.

So, an approximate 95% confidence interval for the gender wage gap is
[Since the estimated gender gap here is simply b~2~, the CI could of
course have also been obtained directly from the **lm** half of the
output.  But with an S having more than two levels, e.g. race, the
**dsld** enhancement is quite valuable.]{.column-margin}

> 13098.2091 &plusmn; 1.96 x 790.4451

or (11548.94,14647.48).

### With-interactions model

As discussed above, in our model

> mean W = $\beta$~0~ + $\beta$~1~ A + $\beta$~2~ M

we identified b~2~ as *the* difference in mean wage between men and
women, regardless of age, so that for instance:

> According to the model, younger men earn about $13,000 more than younger
> women, with the same-sized gap between older men and older women.

But that may not be true.  On the contrary, gender discrimination and
[Interaction between two types of discrimination is called
*intersectionality* by some analysts.]{.column-margin}
age discrimination may interact.  It may be, for instance, that the
gender gap is small at younger ages but much larger for older people.

Technically, the with-interactions model adds a product term:

> mean W = $\beta$~0~ + $\beta$~1~ A + $\beta$~2~ M + $\beta$~3~ AM 

So for example, the gender pay gap for people of age 36 is

($\beta$~0~ + $\beta$~1~ 36 + $\beta$~2~ 1 + $\beta$~3~ 36) -
($\beta$~0~ + $\beta$~1~ 36 ) =

$\beta$~2~ 1 + $\beta$~3~ 36

And at age 43, the gap is

$\beta$~2~ 1 + $\beta$~3~ 43

So, this model does indeed allow for interaction between age and gender.

However, this added-product-term is a bit abstract, and it is easier
(and approximately equivalent) to simply fit two linear models, one for
men and one for women.

::: {.callout-note}
The **dsldLinear** function includes an argument **interactions**. The
default value is FALSE, but if TRUE, it fits separate linear models for
each level of S.  An additional argument 'newData' is now
needed, through which the user specifies a data frame consisting of
one or more (X,C) values at which to compare the effect of S.
:::

```{r}
newData <- data.frame(age=c(36,43))
z <- dsldLinear(svcensus1,'wageinc','gender',interactions=T,
   newData)
summary(z)
```

So the gender pay gap is estimated to be -13156.88 at age 36, -13039.27
at age 43, only about $100.  The estimated gap between ages 36 and 53,
not shown, is larger, close to $300, but it seems there is not much
interaction here.

### Linearity and other assumptions

As noted, linear models are ubiquitous in observational data analysis.
Open any professional journal in medicine, sociology, economics and so
on, and you'll see many applications of this methodology.  But how would one
check that most basic assumption, the linearity of the mean Y for given
X, C and S values?

::: {.callout-note}
### Assumptions--not just a formality

Assumptions *matter*.  They are never perfectly satisfied, but failure
to be even approximately valid can mean deciding that there is no
discrimination when it actually is there, or vice versa.  It can mean
bad medication being declared by the government as good, or vice versa.
In litigation, if a key expert witness is exposed by opposing counsel as
not having checked the assumptions in his/her analysis, the side for
which the witness was testifying will likely lose the case on the spot.

:::

Typically, linearity is checked graphically.  A common approach involves
plotting the *residuals*, which are the differences between the fitted
line and the Y values.  Here, though, we use another graphical approach,
via a **dsld** function that may be more informative. 

Returning to our earlier setting with just S = gender for our example,
we run

```{r}
dsldConditDisparity(svcensus1,'wageinc','gender',
   'age','age > 0',yLim=c(0,150000)) 
```

The function plots a smoothed graph of Y against a user-specified C
[The function has a 'conditions' argument; we have none here, so we just used a
trival one, 'age > 0']{.column-margin} variable, once for each level of
S.  So, the call here says, "Plot wage income against age, for each
gender."  

The model has mean Y being a linear of function of age, so we  should
expect to see approximate straight lines here.  Yet the relation
certainly looks nonlinear, possibly reflecting age discrimination
against both very young and very old workers.  We are already
investigating one kind of discrimination here, gender, so again for
simplicity let's just keep age as a confounder.  

### Updated model

But we must do something about the substantial nonlinearity [Adding a
squared term does not make our model nonlinear, as it is still linear in
the $\beta$~i~; if we, say, double each of those, the entire expression is
doubled, the definition of linearity. The model is nonlinear in age but
linear in the $\beta$~i~.]{.column-margin} we've discovered,
and one possible remedy is to add an age^2^ term be added to the
equation:

mean W = $\beta$~0~ + $\beta$~1~ A + $\beta$~2~ A^2^ + $\beta$~3~ M

Let's fit the updated model:

```{r}
svcensus1$age2 <- svcensus1$age^2
z <- dsldLinear(svcensus1,'wageinc','gender')
coef(z)  # print the estimated coefficients b_i 
```

So we see that the original wage gap figure of about $13,000 was
incorrect, underestimating it by about 15%.

We see in this example that misspecifying a linear model can have a
major impact on its accuracy.  Further issues of model assumptions are
beyond the scope of this book, but the interested reader is referred one
of the most popular applied linear models books, *Regression Modeling
Strategies: With Applications to Linear Models, Logistic and Ordinal
Regression, and Survival Analysis*, by Prof. Frank Harrell, Jr. of the
Vanderbilt University School of Medicine.

### Other assumptions

Other than linearity, the standard errors reported by **lm()** also [It
is also assumed that wage income has a normal/gaussian distribution at
each level, but the Central Limit Theorem's implications for the sums
created by **lm()** are in fact approximately normal.]{.column-margin}
assume that variance of wage income is approximately constant across
ages and genders.  Lack of this property has some effect on the accuracy
of reported standard errors, but this can be adjusted via the so-called
*sandwich* operation, an option in **dsldLinear()**.

## S may consist of more than one factor 

In introducing this example, we noted the need to start simple.  Let's
move away from that a bit.

In the **svcensus** data, both age and gender are potential areas of
discrimination.  We can treat both as such by combining these two R
factor variables into one "super R factor," as follows.

We'll need to discretize age, and since federal law on age
discrimination uses age 40 as the definition of "older," let's use that
as an example:

```{r}
age <- svcensus$age
age <- ifelse(age >= 40,'40+','under40')
age <- as.factor(age)
head(age)
svcensus$age <- age
```

Now let's make our "super factor," using a function from **qeML**:

```{r}
AgeGend <- cartesianFactor('svcensus',c('age','gender')) 
svcensus$AgeGend <- AgeGend
head(svcensus)
```

We no longer need the original age and gender columns, so we'll delete
them and then try some analysis:

```{r}
svcensus$age <- NULL
svcensus$gender <- NULL
w <- dsldLinear(svcensus,'wageinc','AgeGend')
summary(w)
```

Ah, this is a more nuanced probe than the ones above in which we simply
[This is an example of *Simpson's Paradox*, in which an overall average
effect might be large than the terms that make up the average.]{.column-margin}
used age as a confounder.  The male-female differences at both the older
and younger age levels, about $9800 and $7800, are both substantial, but
smaller than the $13,100 overall figure we obtained earlier.  

Note too the impact of age within genders.  Older women made about $8200
more than younger women, but for men the figure was rather larger, about
$10,200.

On the other hand, this analysis is probably too coarse with respect to
age, as it does not reveal the negative impact of age well beyond 40.
It may be worth trying a finer discretization of age, say, 35-, 
35-55 and 55+.

## Logistic model example: German credit data

The *logistic* model is an example of a *generalized linear model*,
whose name stems from it having a linear component in the formula.

### General form of the model

Just as linear models are the most commonly used for numeric Y, in the
binary-Y case the go-to standard is the logistic model.  To introduce
it, let's first consider a very simple prediction problem, in which Y is
gender, say 1 for male, 0 for female, and X is simply income, using the
**svcensus** data.

::: {.callout-note}
### Probability is a special case of a mean
Note that mean Y is now the probability that Y = 1.  That's because the
average of a bunch of 1s and 0s is the proportion of 1s.  In the data
1,0,1,1 for instance, the mean is (1+0+1+1) / 4 = 3/4, and indeed 3/4 of
those numbers are 1s.
:::

Now consider a simple setting in which we are predicting Y = gender from
a single variable X = income (no C here).  Suppose that within each
gender, X has a normal (Gaussian) distribution, the familiar
"bell-shaped" curve, with the same standard deviation for each gender.
Then it turns out that one can show mathematically that

probability male =
$$
\frac
{1}{1 + e^{-(\beta_0 + \beta_1 income)}}
$$

That formula follows the form of the *logistic function*,
$f(t) = 1 / [1 + e^{-t}]$, which has the shape of an S-curve:

```{r}
curve(1/(1+exp(-x)),-4,4)
```

The estimates b~i~ of the population values $\beta_i$ are obtained via a
method generalizing the least-squares method used in the linear case.
[There is a similar situation for the linear case.  If Y and the
predictor variables have a multivariate normal distribution, one can
show that mean Y as a function of the features is linear in the
features, etc.]{.column-margin} Say we predict gender from age and wage
income.  If the latter two variables have a *bivariate normal*
distribution (two-dimensional histogram, has a 3-D bell shape) with the
same variance matrices within each gender, it turns out that we again
get a logistic form:

probability male =
$$
\frac
{1}{1 + e^{-(\beta_0 + \beta_1 age + \beta_2 income)}}
$$

Now, what about that assumption of the normal distributions and so on?
Just as many regression functions for numeric Y in practice are roughly
linear, in predicting binary Y the S-curve model is often roughly valid.
Moreover, the logistic model has two desirable properties for predicting
binary Y:

* Its value is between 0 and 1, appropriate for modeling a probability.

* The expression $1 / [1 + e^{-t}]$ is increasing in t, which we wish to
model when our predictors have monotonic relations with Y.

The point is that the logistic (popularly referred to as "logit") is
[Some books motivate the logistic approach as the *log-odds* ratio,
meaning in this example that the logarithm of the ratio (probability
male) / (probability female) is linear in age and income.]{.column-margin} 
often a good model for binary-Y settings in general.

### We no longer have a no-interactions case

In the first linear model example above, predicting wage income from age
and gender, the model implied a uniform gender gap, independent of age.  The
male and female regression lines were parallel.  That's not possible in
the logit case:

```{r}
curve(1/(1+exp(-x)),-4,4)
curve(1/(1+exp(-2*x)),-4,4,add=TRUE)
```

The curves are no longer parallel; they even cross.  They typically
don't cross in applied settings, but they are always nonparallel.

The implication of this is the same as in the linear case.  We cannot
speak of "the" impact of S on Y, as it will be the same at different
levels of the X and C variables.  Our calls to the software will then
need to specify our desired levels.


### German credit data

This is a built-in dataset to the **fairml** package, which is included
by **dsld**.  Let's take a brief look at the data.

```{r}
data(german.credit)
gct <- german.credit
head(gct)
names(gct)
```

Column 20 of the data, **Credit_risk**, is the assessment made by the
lender, recorded as 'GOOD' or 'BAD'.  Is the lender prejudiced against
women?  We might try to investigate that via a logistic model in which Y
is credit risk, S is gender and we include a few possible confounders.
(Again, let's keep things simple in this example.)

```{r}
gct <- gct[,c('Credit_history','Credit_amount','Present_employment_since',
   'Credit_risk','Gender')]
compData <- gct[c(2,22,222,3,33,55,555),1:3]
w <- dsldLogit(gct,'Credit_risk','Gender',compData,yesYVal='GOOD')
summary(w)
```

As noted, there is no no-interaction case in logit settings, so we need
to specify one or more examples at which we wish to gauge the impact, if
any, of S on Y.  Here we are using rows 2-4 of our data (excluding Y and S)
for that purpose. 

Also, we need to tell the software what values of the credit risk
variable should be consider Y = 1 and Y = 0; we use the **yesYVal**
argument for that, saying we want 'GOOD' to play the Y = 1 role.

Women seem to be faring better than, or at least equal to, men.  If
anything, it is the men who are discrimination victims.  But again, the
size of the gender gap seems to depend a lot on the other factors.
Let's take a look at those comparison cases:

```{r}
compData
```

No salient pattern here, but there is a suggestion that among those who
have been with their present employer, women do better than men.  Not an
easy question at all, and further investigation would be needed.  One
key point: 

::: {.callout-note}
### Role of the software
The software cannot make your decisions for you.  It merely plays a
supporting, and there are no neat formulas to follow.
:::

## Machine Learning Approaches

We referred to the quantities $\beta_i$ above as "parameters."  The
linear and logistic models are thus called *parametric* models.  In each
case, the regression function is modeled as linear, "S curve-shaped" and
so on, that can be described with just a few quantities.

But these models make assumptions, such as assuming the regression
function is approximately linear.  It would also be nice to conduct
analyses that don't rely on such assumptions.

*Machine learning* (ML) algorithms typically do not assume the regression
function has any particular form.  They are thus "safer," though on the
other hand they are less interpretable than, say, that $\beta_2$
quantity in our no-interactions linear model above. 

ML is mainly concerned with prediction, while we here are interested in
estimation of effect sizes.  However, ML algorithms either directly or
indirectly do their prediction by estimating the regression function, as
we do here, so they can be quite useful.

### The k-Nearest Neighbor algorithm

We'll focus here on the k-NN algorithm, as it is the simplest to
explain.  Say we are predicting people's weights, knowing only their
heights.  We have training data, on which we know both the height and
weight of each person.  Faced with a new case of a person 70 inches
tall, we find the people in our training set whose heights are closest to
70 inches, and average their weights.  That average is our predicted
weight for the new person.

How is "close to" defined?  Should we look at the closest 5 people, the
closest 50, or what?  That number is k, the number of nearest neighbors.
It's chosen by the analyst, just like the analyst chooses the number of
bins in plotting a histogram.  Typically k is chosen by trying several
different values of k, then using the one that best predicts in the
holdout set.


### The Law School Admissions dataset

This is the dataset **law.school.admissions**, includein **dsld** via
**fairml**.  It's quite well-known in the ML world, but its full
provenance is unclear.  For instance, the age variables skews to
middle-aged and older, seemingly not consistent with the data's
description on Kaggle.  Thus it should be kept in mind that this is just
an illustration of methodology, and firm conclusions about the legal
education process should not be drawn.  The data concern students who
were admitted to law school, so in spite of the title, it's not about
the admissions process itself.

The main variables of interest here are:

* **decile1**, **decile2**:  Student's standing after Year 1 and Year 3
of law school.

* **fam_inc**: Apparently the income level of the family in which the
law students was raised in.

* **lsat**: Score on the LSAT, a major law school admissions test.

* **ugpa**:  Undergraduate GPA.

* **gender**:  Gender.

* **race1**:  Primary racial group.

* **cluster**:  Level of prestige of the law school.

* **bar**:  After law school, did this person pass the bar examination?

There has been concern that the LSAT is heavily influenced by family
income, thus unfair, especially to Black and Latino students.  Let's
investigate using **dsld**.

```{r}

```

```
